### 2014-05-19 text/Chapter04.tex
Categorical semantics of flat calculus (Chapter 4)

a

\subsection{Comonads}
\subsection{Indexed comonads}
\subsection{Semantics of flat calculus}
\subsection{Examples}

@@

% comonadic semantics
\newcommand{\iftor}[1]{C^{#1}}
\newcommand{\iftorhat}[1]{\hat{C}^{#1}}

\newcommand{\cosem}[1]{\llbracket #1 \rrbracket}
\newcommand{\interp}[1]{\llbracket #1 \rrbracket}
\newcommand{\ccat}[0]{\mathcal{C}}
\newcommand{\obj}[1]{\textnormal{obj}(#1)}

\newcommand{\cobind}[2]{#1^\dagger_{#2}}
\newcommand{\cmerge}[0]{ \ident{m} }
\newcommand{\csplit}[0]{ \ident{n} }
\newcommand{\counit}[0]{ \varepsilon }

@@

The approach of \emph{categorical semantics} interprets terms as
morphisms in some category.  For typed calculi, typing judgments $x_1 : \tau_1 \ldots x_n : \tau_n
\vdash e: \tau$ are usually mapped to morphisms $\interp{\tau_1} \times \ldots \times
\interp{\tau_n} \rightarrow \interp{\tau}$. Moggi showed the semantics of various effectful 
computations can be captured generally using the (\emph{strong}) \emph{monad}
structure~\cite{monad-notions}. Dually, Uustalu and Vene showed that
(\emph{monoidal}) \emph{comonads} capture various kinds of context-dependent 
computation~\cite{comonads-notions}.

We extend Uustalu and Vene's approach to give a semantics for
the coeffect calculus by generalising comonads to
\emph{indexed comonads}. We emphasise semantic intuition and
abbreviate the categorical foundations for space reasons.

\vspace{-1em}
\paragraph{Indexed comonads.}

Uustalu and Vene's approach interprets well-typed terms 
as morphisms $C (\tau_1 \times \ldots
\times \tau_n) \rightarrow \tau$, where $C$ encodes contexts
and has a comonad structure~\cite{comonads-notions}.
Indexed comonads comprise a \emph{family} of object mappings 
$C^r$ indexed by a coeffect $r$ describing the contextual requirements satisfied by
the encoded context. We interpret judgments $C^r (x_1 : \tau_1, \ldots, x_n : \tau_n) \vdash e : \tau$ as
morphisms $C^r (\interp{\tau_1} \times \ldots \times \interp{\tau_n}) \rightarrow \interp{\tau}$.

The indexed comonad structure provides a notion of composition
for computations with different contextual requirements.
%%
\begin{definition}
Given a monoid $(S, \cseq, \cunit)$ with binary operator $\cseq$ and
unit $\cunit$, an \emph{indexed comonad} over a category $\ccat$
comprises a family of object mappings $\iftor r$ where for all $r \in S$ and $A \in \obj{\ccat}$ then 
$\iftor r A \in \obj{\ccat}$ and:
%%
\begin{itemize}
\item a natural transformation $\counit_A : \iftor{\cunit} A \rightarrow A$, called the \emph{counit};
\item a family of mappings $\cobind{(-)}{r, s}$ from morphisms $\iftor
  r A \rightarrow B$ to \\ morphisms $\iftor{r \cseq{} s} A
  \rightarrow \iftor s B$ in $\ccat$, natural in $A,B$, called
  \emph{coextend};
\end{itemize}
%
\noindent
such that for all $f : \iftor r \tau_1 \rightarrow \tau_2$ and $g : \iftor s \tau_2 \rightarrow \tau_3$ 
the following equations hold:
%
\renewcommand{\arraycolsep}{1.5em}
\begin{align*}
\begin{array}{ccc}
\counit \circ \cobind{f}{r, \cunit} = f & 
\cobind{{(\counit)}}{\cunit, r} = \ident{id} & 
\cobind{(g \circ \cobind{f}{r, s})}{(r \cseq s), t} = \cobind{g}{s, t} \circ \cobind{f}{r, (s \cseq t)} \\
\end{array}
\end{align*}
\end{definition}
%
The \emph{coextend} operation gives rise to an associative composition operation for
computations with contextual requirements (with \emph{counit} as the identity):
%
\[
\hat{\circ} : (\iftor r \tau_1 \rightarrow \tau_2) \rightarrow (\iftor s \tau_2 \rightarrow \tau_3) 
  \rightarrow (\iftor{r \cseq s} \tau_1 \rightarrow \tau_3) \quad\quad\quad
g \, \hat{\circ} \, f = g \circ f^\dagger_{r,s}\quad
\]
%
The composition $\hat{\circ}$ best expresses the intention
of indexed comonads: contextual requirements of the composed
functions are combined. The properties of
the composition follow from the indexed comonad laws and the 
monoid $(S, \oplus, \cunit)$.

\paragraph{Example}
Indexed comonad are analogous to comonads (in coKleisli form), but
with the additional monoidal structure on indices.  Indeed, comonads
are a special case of indexed comonads with a trivial singleton
monoid, \eg{}, $(\{1\}, \ast, 1)$ with $1 \ast 1 = 1$ where $\iftor 1$
is the underlying functor of the comonad and $\counit$ and
$\cobind{(-)}{1, 1}$ are the usual comonad operations.
However, as demonstrated next, not all indexed comonads are derived from ordinary comonads.

\paragraph{Example}
The \emph{indexed partiality comonad} encodes free-variable contexts
of a computation which are either \emph{live} or \emph{dead} (\ie{}, have
\emph{liveness} coeffects) with the monoid $(\{\ident{D}, \ident{L}\},
\sqcap, \ident{L})$, where $\iftor{\ident L} A = A$ encodes live
contexts and $\iftor{\ident D} A = 1$ encodes dead contexts,
where $1$ is the unit type inhabited by a single value $()$.
The \emph{counit} operation $\counit{} : \iftor{\ident{L}} A \rightarrow A$
is defined $\counit\ x = x$ and \emph{coextend},
for all $f : \iftor r A \rightarrow B$, and thus
$\cobind{f}{r, s} : \; \iftor{r \sqcap s} A \rightarrow \iftor s B$,
is defined:
%
\begin{align*}
\begin{array}{llll}
\cobind{f}{\ident{D}, \ident{D}} x = () \qquad\qquad & 
\cobind{f}{\ident{D}, \ident{L}} x = f () \qquad\qquad &
\cobind{f}{\ident{L}, \ident{D}} x = () \qquad\qquad &
\cobind{f}{\ident{L}, \ident{L}} x = f~x 
\end{array}
\end{align*}
%
The indexed family $C^r$ here is analogous to the non-indexed \textsf{Maybe} (or \emph{option})
data type $\mathsf{Maybe}\ A = A + 1$.  This
type does not permit a comonad structure since
$\varepsilon : \mathsf{Maybe}\ A \rightarrow A$ is undefined at $(\textsf{inr} \, ())$. 
For the indexed comonad, $\counit$ need only be defined
 for $\iftor{\ident{L}} A = A$. Thus, indexed comonads capture a broader range of
 contextual notions of computation than comonads.

 Moreover, indexed comonads are not restricted by the \emph{shape
   preservation} property of comonads~\cite{orchard12codo}: that
 a coextended function cannot change the \emph{shape} of the
 context. For example, in the second case above $f^\dagger_{\ident{D},
   \ident{L}} : \iftor{\ident{D}} A \rightarrow \iftor{\ident{L}} B$
 where the shape changes from $1$ (empty context) to $B$ (available
 context).

@@

The semantics of the coeffect calculus requires an indexed
lax semi-monoidal structure for combining contexts \emph{as well as} an indexed
\emph{colax} monoidal structure for \emph{splitting} contexts. These are provided
by two families of morphisms (given a coeffect algebra with $\vee$ and $\wedge$):
%
\begin{itemize}
\item $\cmerge_{r, s} : \iftor r A \times \iftor s B \rightarrow \iftor{(r \wedge s)}(A \times B)$ natural
in $A, B$;
\item $\csplit_{r, s} : \iftor{(r \vee s)}(A \times B) \rightarrow \iftor r A \times \iftor s B$ natural
in $A, B$;
\end{itemize}
%
\noindent
The $\cmerge_{r, s}$ operation merges contextual computations with
tags combined by $\wedge$ (greatest lower-bound), elucidating 
the behaviour of $\cmerge_{r, s}$: that merging may result in 
the loss of some parts of the contexts $r$ and $s$.

@@

%\begin{example}
%The semantics of the coeffect calculus for
% dynamically-scoped implicit parameters, with coeffect
%algebra $(\mathcal{P}(\ident{Id}), \cup, \cup, \cup, \emptyset)$, 
%is provided by the indexed comonad $\iftor rA = A \times (r \rightarrow \tau)$
%where $r \in \mathcal{P}(\ident{Id})$ with operations:\dnote{The type here is wrong,
%do we need to store the types too? Maybe should just be a product.}
%%
%\vspace{-1.0em}
%\begin{equation*}
%\begin{array}{l}
%\cobind{f}{r, s} \; : \; \iftor{r \cup s} A \rightarrow \iftor s B\\
%\cobind{f}{r, s} (a, g) = (f(a, g|_r), g|_s)
%\\[1em]
%\counit{} (a, g) = a\\
%%\iota_{r, s} \; : \: \iftor r A \rightarrow \iftor s A\\
%\iota_{r, s}(a, g) = (a, g|_s)\quad(\textnormal{if}~\leq r)
%\end{array}
%\quad\begin{array}{l}
%\cmerge_{r, s} : \iftor r A \times \iftor s B \rightarrow \iftor{(r \cup s)}(A \times B)\\
%\cmerge_{r, s} ((a_1, g_1), (a_2, g_2)) = ((a_1, a_2), g_1|_{r \setminus s} \cup g_2 )
%\\[1em]
%\csplit_{r, s} : \iftor{(r \cup s)}(A \times B) \rightarrow \iftor r A \times \iftor s B \\
%\csplit_{r, s} ((a_1, a_2), g) = ((a_1, g|_r), (a_1, g|_s))
%\end{array}
%\end{equation*}
%\end{example}

@@

\begin{figure*}[t]
\vspace{-1em}
\newcommand{\uu}[1]{\overline{#1}}
% \begin{equation*} 
% \cosem{\ctyp{r}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
%   \ctyp{r}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
% \end{equation*}
% \vspace{-0.4em}
\begin{equation*}
\begin{array}{rcll}
\cosem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
    \emph{curry} \; (\cosem{\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    %&$(1)$
\\[0.1em]
\cosem{\ctyp{r \vee (s \cseq t)}{\Gamma} \vdash e_1~e_2 : \tau} &=& 
    (\emph{uncurry} \;  \cosem{\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2}) \, \circ \\
 & & \quad\; (id \times \cobind{\cosem{\ctyp{t}{\Gamma}\vdash e_2 : \tau_1}}{t,s})
    \circ \csplit_{r, s \cseq t} \circ \iftor{r \vee (s \cseq t)}\Delta
\\[0.1em]
\cosem{\ctyp{\cunit}{\Gamma} \vdash x_i : \tau_i} &=&
    \pi_i \circ \counit                                                                  % &$(3)$
%\\[0.1em]
%\cosem{\ctyp{r}{\Gamma} \vdash e : \tau} &=&
%    \cosem{\ctyp{s}{\Gamma} \vdash e : \tau} \circ \iota_{r, s}                              %   &%$(4)$
%\\[0.1em]
%\cosem{\ctyp{r\vee (r \cseq s)}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2} &=&
%    \cosem{\ctyp{r}{(\Gamma, x:\tau_1)} \vdash e_2 : \tau_2} \circ
% ((\varepsilon \times \cosem{ \ctyp{s}{\Gamma} \vdash e_1 : \tau_1 }) \circ 
%\iftorhat{r \vee (r \cseq t)}\Delta
\end{array}
\end{equation*}
\vspace{-1.5em}
\caption{Categorical semantics for the coeffect calculus}
\label{fig:semantics}
\vspace{-1.0em}
\end{figure*}

>>>>

The type system of flat coeffect calculus arises as a generalization of the examples discussed in 
Chapter~\ref{ch:applications}, but we can also obtain it by looking at the categorical semantics
of context-dependent computations. This is a direction that we explore in this section. Although
the development presented here is interesting in its own, our main focus is \emph{using} categorical
semantics to motivate and explain the design of flat coeffect calculus.

\subsection{Categorical semantics}

As discussed in Section~\ref{sec:path-sem}, categorical semantics interprets terms as morphisms
in some category. For typed calculi, the semantics defined by $\sem{-}$ usually interprets typing 
judgements $x_1 \!:\! \tau_1 \ldots x_n \!:\! \tau_n \vdash e: \tau$ as morphisms 
$\sem{\tau_1} \times \ldots \times \sem{\tau_n} \rightarrow \sem{\tau}$.

As a best known example, Moggi \cite{monad-notions} showed that the semantics of various effectful 
computations can be captured uniformly using the (\emph{strong}) \emph{monad} structure. In that 
approach, computations are interpreted as $\tau_1 \times \ldots \times \tau_n \rightarrow \mtyp{}{\tau}$
for some monad $\mtyp{}{}$. For example, $\mtyp{}{\alpha} = \alpha \cup \{ \bot \}$ models 
partiality (maybe monad), $\mtyp{}{\alpha} = \mathcal{P}(\alpha)$ models non-determinism (list 
monad) and $\mtyp{}{\alpha} = (\alpha \times S)^S$ models side-effects (state monad). Here, the 
structure of a strong monad provides necessary ``plumbing'' for composing monadic computations.

Following similar approach to Moggi, Uustalu and Vene \cite{comonads-notions} showed that 
(\emph{monoidal}) \emph{comonads} uniformly capture the semantics of various kinds of context-dependent 
computations~\cite{comonads-notions}. For example, data-flow computations over non-empty lists
$\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$ are modelled using
the non-empty list comonad.

The monadic and comonadic model outlined here represents at most a binary analysis of effects or 
context-dependence. A function $\tau_1 \rightarrow \tau_2$ performs \emph{no} effects (requires no 
context) whereas $\tau_1 \rightarrow \mtyp{}{\tau_2}$ performs \emph{some} effects and
$\ctyp{}{\tau_1} \rightarrow \tau_2$ requires \emph{some} context. In the next section, we introduce
\emph{indexed comonads}, which provide a more precise analysis and let us model computations with
context requirements $\cclrd{r}$ as functions $\ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2$ using
an \emph{indexed comonad} $\ctyp{\cclrd{r}}{}$.

\subsection{Introducing comonads}

In category theory, \emph{comonad} is a dual of \emph{monad}. Informally, we get a comonad by 
taking a monad and ``reversing the arrows''. More formally, one of the equivalent definitions of
comonad looks as follows:

\begin{definition}
A \emph{comonad} over a category $\catc$ is a triple $(C, \ident{counit}, \ident{cobind})$ where:
\begin{compactitem}
\item $C$ is a mapping on objects (types) $C : \catc \rightarrow \catc$
\item $\ident{counit}$ is a mapping $\ctyp{}{\alpha} \rightarrow \alpha$ 
\item $\ident{cobind}$ is a mapping $(\ctyp{}{\alpha} \rightarrow \beta) 
  \rightarrow (\ctyp{}{\alpha} \rightarrow \ctyp{}{\beta})$
\end{compactitem}
such that, for all $f:\ctyp{}{\alpha} \rightarrow \beta$ and $g:\ctyp{}{\beta} \rightarrow \gamma$:
\begin{align}
\tag{\emph{left identity}}
  \ident{cobind}~\ident{counit} &= \idf{}
  \\
\tag{\emph{right identity}}
  \ident{counit} \circ \ident{cobind}~f &= f
  \\
\tag{\emph{associativity}}
  \ident{cobind}~(g \circ \ident{cobind}~f) &= (\ident{cobind}~g) \circ (\ident{cobind}~f)
\end{align}
\end{definition}

\noindent
From the functional programming perspective, we can see $\ctyp{}{}$ as a parametric data type such as
\ident{NEList}. The $\ident{counit}$ operations extracts a value $\alpha$ from a value that carries 
additional context $\ctyp{}{\alpha}$. The $\ident{cobind}$ operation turns a context-dependent function 
$\ctyp{}{\alpha} \rightarrow \beta$ into a function that takes a value with context, applies
the context-dependent function to value(s) in the context and then propagates the context.

As mentioned earlier, Uustalu and Vene \cite{comonads-notions} use comonads to model data-flow
computations. They describe infinite (coinductive) streams and non-empty lists as example comonads.

\begin{example}[Non-empty list]
A non-empty list is a recursive data-type defined as $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
We write \kvd{inl} and \kvd{inr} for constructors of the left and right cases, respectively. The 
type \ident{NEList} forms a comonad together with the following \ident{counit} and \ident{cobind} mappings:
%
\begin{equation*}
\begin{array}{rclll}
\ident{counit}~l &\narrow{=}& h &\quad&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
\ident{counit}~l &\narrow{=}& h &&\textnormal{when}~l=\kvd{inr}~(h, t)\\[0.5em]
\ident{cobind}\,f~l &\narrow{=}& \kvd{inl}~(f\,l) &&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
\ident{cobind}\,f~l &\narrow{=}& \kvd{inr}~(f\,l,\;\ident{cobind}~f~t) &&\textnormal{when}~l=\kvd{inr}~(h, t)
\end{array}
\end{equation*}
\end{example}

\noindent
The \ident{counit} operation returns the head of the non-empty list. Note that it is crucial that
the list is \emph{non-empty}, because we always need to be able to obtain a value. The \ident{cobind}
defined here returns a list of the same length as the original where, for each element, the 
function $f$ is applied on a \emph{suffix} list starting from the element. Using a simplified
notation for list, the result of applying \ident{cobind} to a function that sums elements of a
list gives the following behaviour:
%
\begin{equation*}
\ident{cobind}~\ident{sum}~(7,6,5,4,3,2,1,0) = (28,21,15,10,6,3,1,0)
\end{equation*}
%
The fact that the function $f$ is applied to a \emph{suffix} is important in order to satisfy the
\emph{left identity} law, which requires that $\ident{cobind}~\ident{counit}~l = l$.

It is also interesting to examine some data types that do \emph{not} form a comonad. As already
mentioned, list $\ident{List}~\alpha = 1 + (\alpha \times \ident{List}~\alpha)$ is not a comonad,
because the \ident{counit} operation is not defined for the value $\kvd{inl}~()$. Similarly,
the \ident{Maybe} data type defined as $1 + \alpha$ is not a comonad for the same reason.
However, if we consider flat coeffect calculus for liveness, it appears natural to model computations
as function $\ident{Maybe}~\tau_1 \rightarrow \tau_2$. To use such model, we first need to 
generalise comonads to \emph{indexed comonads}.

\subsection{Generalising to indexed comonads}
The flat coeffect algebra includes a monoid $(\C, \cseq, \cunit)$, which defines the behaviour of
sequential composition, where the annotation $\cunit$ represents a variable access. An indexed 
comonad is formed by a data type (object mapping) $\ctyp{\cclrd{r}}{\alpha}$ where the annotation 
$\cclrd{r}$ determines what context is required.

\begin{definition}
Given a monoid $(\C, \cseq, \cunit)$ with binary operator $\cseq$ and unit $\cunit$, an 
\emph{indexed comonad} over a category $\catc$ is a triple 
$(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, \ident{cobind}_{\cclrd{r}, \cclrd{s}})$ where:

\begin{compactitem}
\item $\ctyp{\cclrd{r}}{}$ for all $\cclrd{r} \in \C$ is a family of object mappings 
\item $\ident{counit}_{\cunit}$ is a mapping $\ctyp{\cunit}{\alpha} \rightarrow \alpha$ 
\item $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$ is a mapping $(\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
  \rightarrow (\ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta})$
\end{compactitem}
such that, for all $f:\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta$ and $g:\ctyp{\cclrd{s}}{\beta} \rightarrow \gamma$
and the identity $\idf{\cclrd{s}} : \ctyp{\cclrd{s}}\alpha \rightarrow \ctyp{\cclrd{s}}\alpha$:
\begin{align}
\tag{\emph{left identity}}
  \ident{cobind}_{\cunit, \cclrd{s}}~\ident{counit}_{\cunit} &= \idf{}
  \\
\tag{\emph{right identity}}
  \ident{counit}_{\cunit} \circ \ident{cobind}_{\cclrd{r}, \cunit}~f &= f
  \\
\tag{\emph{associativity}}
\hspace{-10em}
  \ident{cobind}_{\cclrd{r}\cseq\cclrd{s},\cclrd{t}}~(g \circ \ident{cobind}_{\cclrd{r}, \cclrd{s}}~f) &= 
    (\ident{cobind}_{\cclrd{s}, \cclrd{t}}~g) \circ (\ident{cobind}_{\cclrd{r}, \cclrd{s}\cseq\cclrd{t}}~f)
\end{align}
\end{definition}

\noindent
Rather than defining a single mapping $\ctyp{}{}$, we are now defining a family of mappings 
$\ctyp{\cclrd{r}}{}$ indexed by a monoid structure. Similarly, the operation $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$ 
operation is now also formed by a \emph{family} of mappings for different pairs of indices 
$\cclrd{r}, \cclrd{s}$. To be fully precise, $\ident{cobind}$ is a family of natural transformations 
and we should include $\alpha, \beta$ as indices, writing $\ident{cobind}_{\cclrd{r},\cclrd{s}}^{\alpha, \beta}$.
For the purpose of this thesis, it is sufficient to treat $\ident{cobind}$ as a family of
mappings or, when it does not  introduce ambiguity, view it as a single mapping.

The $\ident{counit}$ operation is not defined for all $\cclrd{r} \in \C$, but only for 
the unit $\cunit$. We still include the unit as an index writing $\ident{counit}_{\cunit}$, 
but this is merely for symmetry. Crucially, this means that the operation is defined only
for some special contexts.

If we look at the indices in the laws, we can see that the left and right identity 
require $\cunit$ to be the unit of $\cseq$. Similarly, the associativity law implies the 
associativity of the $\cseq$ operator.

The category that models sequential composition is formed by the unit arrow $\ident{counit}$ 
together with the (associative) composition operation that composes computations with 
contextual requirements as follows:
%
\begin{equation*}
\begin{array}{ccl}
\textnormal{--}\, \hat{\circ} \,\textnormal{--}&\narrow{:}& (\ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2) 
  \rightarrow (\ctyp{\cclrd{s}}{\tau_2} \rightarrow \tau_3) 
  \rightarrow (\ctyp{\cclrd{r} \cseq \cclrd{s}}{\tau_1} \rightarrow \tau_3) \\
g \, \hat{\circ} \, f &\narrow{=}& g \circ (\ident{cobind}_{\cclrd{r}, \cclrd{s}} f)
\end{array}
\end{equation*}
%
The composition $\hat{\circ}$ best expresses the intention of indexed comonads. Given two functions
with contextual requirements $\cclrd{r}$ and $\cclrd{s}$, their composition is a function that 
requires $\cclrd{r}\,\cseq,\cclrd{s}$. The contextual requirements propagate \emph{backwards} and
are attached to the input of the composed function.

\subsection{Examples}

Any comonad can be turned into an indexed comonad using a trivial monoid. However, indexed comonads
are more general and can be used with other data types, including indexed \ident{Maybe}.

\begin{example}[Comonads]
Any comonad $\ctyp{}{}$ is an indexed comonad with an index provided by a trivial monoid $(\{1\},\ast,1)$
where $1\ast 1 = 1$ and $\ctyp{1}{}$ is the underlying mapping $\ctyp{}{}$ of the original comonad. The
operations $\ident{counit}_1$ and $\ident{cobind}_{1,1}$ are defined by the operations $\ident{counit}$
and $\ident{cobind}$ of the comonad.
\end{example}

\begin{example}[Indexed option]
The indexed option comonad is defined over a monoid $(\{ \ident{L},\ident{D} \}, \sqcup,\ident{L})$ 
where $\sqcup$ is defined as earlier, \ie~$\ident{L} = \cclrd{r} \sqcup \cclrd{s} \Longleftrightarrow \cclrd{r}=\cclrd{s}=\ident{L}$.
Assuming $1$ is the unit type inhabited by $()$, the mappings are defined as follows:
\begin{equation*}
\begin{array}{l}
\ctyp{\ident{L}}{\alpha} = \alpha\\
\ctyp{\ident{D}}{\alpha} = 1\\
\\
\ident{counit}_{\ident{L}} : \ctyp{\ident{L}}{\alpha} \rightarrow \alpha \\
\ident{counit}_{\ident{L}}~v = v\\
\end{array}
\qquad
\begin{array}{lcl}
\multicolumn{3}{l}{
  \ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
    \rightarrow (\ctyp{\cclrd{r}\sqcup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) }\\
\ident{cobind}_{\ident{L},\ident{L}}~f~x &\narrow{=}& f~x\\
\ident{cobind}_{\ident{L},\ident{D}}~f~() &\narrow{=}& ()\\
\ident{cobind}_{\ident{D},\ident{L}}~f~() &\narrow{=}& f~()\\
\ident{cobind}_{\ident{D},\ident{D}}~f~() &\narrow{=}& ()\\
\end{array}
\end{equation*}
\end{example}

\noindent
The indexed option comonad models the semantics of the liveness coeffect system discussed in 
\ref{sec:applications-flat-live}, where $\ctyp{\ident{L}}{\alpha} = \alpha$ models a live context 
and $\ctyp{\ident{D}}{\alpha}=1$ models a dead context which does not contain a value. The \ident{counit}
operation extracts a value from a live context; \ident{cobind} can be seen as an implementation of 
dead code elimination. The definition only evaluates $f$ when the result is marked as live and is thus
required, and it only accesses $x$ if the function $f$ requires its input.

The indexed family $\ctyp{\cclrd{r}}{}$ in the above example is analogous to the \ident{Maybe}
(or option) data type $\ident{Maybe}\,\alpha = 1 + \alpha$. As mentioned earlier, this type does not 
permit (non-indexed) comonad structure, because $\ident{counit}~()$ is not defined. This is not a 
problem with indexed comonads, because \ident{counit} only needs to be defined on live context.

\begin{example}[Indexed product]
The semantics of implicit parameters

\end{example}

\paragraph{Related work}

Not a family of comonads

Katsumata

\newcommand{\ccat}[0]{\mathcal{C}}
\newcommand{\cobind}[2]{#1^\dagger_{#2}}
\newcommand{\cmerge}[0]{ \ident{m} }
\newcommand{\csplit}[0]{ \ident{n} }
\newcommand{\counit}[0]{ \varepsilon }

\paragraph{Example}
Indexed comonad are analogous to comonads (in coKleisli form), but
with the additional monoidal structure on indices.  Indeed, comonads
are a special case of indexed comonads with a trivial singleton
monoid, \eg{}, $(\{1\}, \ast, 1)$ with $1 \ast 1 = 1$ where $\ccat^1$
is the underlying functor of the comonad and $\counit$ and
$\cobind{(-)}{1, 1}$ are the usual comonad operations.
However, as demonstrated next, not all indexed comonads are derived from ordinary comonads.

\paragraph{Example}
The \emph{indexed partiality comonad} encodes free-variable contexts
of a computation which are either \emph{live} or \emph{dead} (\ie{}, have
\emph{liveness} coeffects) with the monoid $(\{\ident{D}, \ident{L}\},
\sqcap, \ident{L})$, where $\ccat^{\ident L} A = A$ encodes live
contexts and $\ccat^{\ident D} A = 1$ encodes dead contexts,
where $1$ is the unit type inhabited by a single value $()$.
The \emph{counit} operation $\counit{} : \ccat^{\ident{L}} A \rightarrow A$
is defined $\counit\ x = x$ and \emph{coextend},
for all $f : \ccat^r A \rightarrow B$, and thus
$\cobind{f}{r, s} : \; \ccat^{r \sqcap s} A \rightarrow \ccat^s B$,
is defined:
%
\begin{align*}
\begin{array}{llll}
\cobind{f}{\ident{D}, \ident{D}} x = () \qquad\qquad & 
\cobind{f}{\ident{D}, \ident{L}} x = f () \qquad\qquad &
\cobind{f}{\ident{L}, \ident{D}} x = () \qquad\qquad &
\cobind{f}{\ident{L}, \ident{L}} x = f~x 
\end{array}
\end{align*}
%
The indexed family $C^r$ here is analogous to the non-indexed \textsf{Maybe} (or \emph{option})
data type $\mathsf{Maybe}\ A = A + 1$.  This
type does not permit a comonad structure since
$\varepsilon : \mathsf{Maybe}\ A \rightarrow A$ is undefined at $(\textsf{inr} \, ())$. 
For the indexed comonad, $\counit$ need only be defined
 for $\ccat^{\ident{L}} A = A$. Thus, indexed comonads capture a broader range of
 contextual notions of computation than comonads.

 Moreover, indexed comonads are not restricted by the \emph{shape
   preservation} property of comonads~\cite{orchard12codo}: that
 a coextended function cannot change the \emph{shape} of the
 context. For example, in the second case above $f^\dagger_{\ident{D},
   \ident{L}} : \ccat^{\ident{D}} A \rightarrow \ccat^{\ident{L}} B$
 where the shape changes from $1$ (empty context) to $B$ (available
 context). 
a

\subsection{Indexed monoidal comonads}

!!!

not a family of comonads

\subsection{Semantics of flat calculus}

@@

@@

@@

The semantics of the coeffect calculus requires an indexed
lax semi-monoidal structure for combining contexts \emph{as well as} an indexed
\emph{colax} monoidal structure for \emph{splitting} contexts. These are provided
by two families of morphisms (given a coeffect algebra with $\vee$ and $\wedge$):
%
\begin{itemize}
\item $\cmerge_{r, s} : \ccat^r A \times \ccat^s B \rightarrow \ccat^{(r \wedge s)}(A \times B)$ natural
in $A, B$;
\item $\csplit_{r, s} : \ccat^{(r \vee s)}(A \times B) \rightarrow \ccat^r A \times \ccat^s B$ natural
in $A, B$;
\end{itemize}
%
\noindent
The $\cmerge_{r, s}$ operation merges contextual computations with
tags combined by $\wedge$ (greatest lower-bound), elucidating 
the behaviour of $\cmerge_{r, s}$: that merging may result in 
the loss of some parts of the contexts $r$ and $s$.

@@

%\begin{example}
%The semantics of the coeffect calculus for
% dynamically-scoped implicit parameters, with coeffect
%algebra $(\mathcal{P}(\ident{Id}), \cup, \cup, \cup, \emptyset)$, 
%is provided by the indexed comonad $\ccat^rA = A \times (r \rightarrow \tau)$
%where $r \in \mathcal{P}(\ident{Id})$ with operations:\dnote{The type here is wrong,
%do we need to store the types too? Maybe should just be a product.}
%%
%\vspace{-1.0em}
%\begin{equation*}
%\begin{array}{l}
%\cobind{f}{r, s} \; : \; \ccat^{r \cup s} A \rightarrow \ccat^s B\\
%\cobind{f}{r, s} (a, g) = (f(a, g|_r), g|_s)
%\\[1em]
%\counit{} (a, g) = a\\
%%\iota_{r, s} \; : \: \ccat^r A \rightarrow \ccat^s A\\
%\iota_{r, s}(a, g) = (a, g|_s)\quad(\textnormal{if}~\leq r)
%\end{array}
%\quad\begin{array}{l}
%\cmerge_{r, s} : \ccat^r A \times \ccat^s B \rightarrow \ccat^{(r \cup s)}(A \times B)\\
%\cmerge_{r, s} ((a_1, g_1), (a_2, g_2)) = ((a_1, a_2), g_1|_{r \setminus s} \cup g_2 )
%\\[1em]
%\csplit_{r, s} : \ccat^{(r \cup s)}(A \times B) \rightarrow \ccat^r A \times \ccat^s B \\
%\csplit_{r, s} ((a_1, a_2), g) = ((a_1, g|_r), (a_1, g|_s))
%\end{array}
%\end{equation*}
%\end{example}

@@

\begin{figure*}[t]
\vspace{-1em}
\newcommand{\uu}[1]{\overline{#1}}
% \begin{equation*} 
% \sem{\ctyp{r}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
%   \ctyp{r}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
% \end{equation*}
% \vspace{-0.4em}
\begin{equation*}
\begin{array}{rcll}
\sem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
    \emph{curry} \; (\sem{\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    %&$(1)$
\\[0.1em]
\sem{\ctyp{r \vee (s \cseq t)}{\Gamma} \vdash e_1~e_2 : \tau} &=& 
    (\emph{uncurry} \;  \sem{\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2}) \, \circ \\
 & & \quad\; (id \times \cobind{\sem{\ctyp{t}{\Gamma}\vdash e_2 : \tau_1}}{t,s})
    \circ \csplit_{r, s \cseq t} \circ \ccat^{r \vee (s \cseq t)}\Delta
\\[0.1em]
\sem{\ctyp{\cunit}{\Gamma} \vdash x_i : \tau_i} &=&
    \pi_i \circ \counit                                                                  % &$(3)$
%\\[0.1em]
%\sem{\ctyp{r}{\Gamma} \vdash e : \tau} &=&
%    \sem{\ctyp{s}{\Gamma} \vdash e : \tau} \circ \iota_{r, s}                              %   &%$(4)$
%\\[0.1em]
%\sem{\ctyp{r\vee (r \cseq s)}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2} &=&
%    \sem{\ctyp{r}{(\Gamma, x:\tau_1)} \vdash e_2 : \tau_2} \circ
% ((\varepsilon \times \sem{ \ctyp{s}{\Gamma} \vdash e_1 : \tau_1 }) \circ 
%\iftorhat{r \vee (r \cseq t)}\Delta
\end{array}
\end{equation*}
\vspace{-1.5em}
\caption{Categorical semantics for the coeffect calculus}
\label{fig:semantics}
\vspace{-1.0em}
\end{figure*}
